Utilizing Parallel Workers: \\LLVM's Vectorization Plan

### 2024-11-23 paper/outline.tex
paper

\begin{abstract}
Lorem ipsum.
\end{abstract}

@@

\begin{enumerate}
  \item Modern CPUs (SIMD-Registers, SIMD ISAs)
  \item SIMD performance gains
  \item Auto-Vectorization to help programmers utilize SIMD functionality out of the box 
  (explicit vectorization as time consuming especially when supporting multiple architectures)
  \item LLVM as popular, platform-agnostic compiler framework~\cite{10.5555/977395.977673}
  \item auto-vectorization complexity $\rightarrow$ wish for well designed, 
  interoperable system (VPlan) $\rightarrow$ transition from Loop Vectorizer + SLP Vectorizer to 
  Vectorization Plan $\rightarrow$ ongoing 8 year long effort started by Intel
  \item rough paper outline
\end{enumerate}

@@

\section{LLVM's Vectorization Plan}
\label{sec:vplan}
  Overview~\cite{llvmvec,llvmvplan,llvmextloopvec,llvmintrvplan,llvmvplanstate,llvmvplanrv,llvmouterloop,llvmvplanupdate}
\begin{enumerate}
  \item Phase 1: Legality
  \item Phase 2: Planning (Vectorization Plans)
  \begin{enumerate}
    \item Building initial VPlan based on Phase 1 Constraints (Generate one or more up-to-date Code Examples for VPlan similar to~\cite{llvmintrvplan,llvmvplanstate,llvmouterloop})
    $\rightarrow$ Explain different components (e.g.: \texttt{VPBasicBlock}, \texttt{VPRegionBlock}, etc.) and
    correlation between generated VPlan structure, original scalar code. Show some general optimizations.
    \item Modelling Cost. Not yet part of VPlan. (related work? already adopted?~\cite{POHL2020102106})
    \item Optimizing/Transforming VPlans~\cite{llvmvplanupdate}
  \end{enumerate}
  \item Phase 3: Execution - Materialize best VPlan
\end{enumerate}

@@

\section{Related Work}
\label{sec:relatedwork}
\begin{enumerate}
  \item Other auto-vectorization algorithms / possible improvements~\cite{10.5555/3615924.3615932,4336219,llvmveccontrolflow,saarlandrv,10.1145/3303117.3306172}
  \\Some of those should already be (partially) implemented in LLVM
  \item Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc}
  \item Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624}
  \item (Polyhedral compilation techniques~\cite{llvmpolyhedral})
  \item (LLM-based Vectorization~\cite{taneja2024llmvectorizerllmbasedverifiedloop})
\end{enumerate}

@@

\section{Summary and Future Work}
\label{sec:summary}
\begin{enumerate}
  \item VPlan Summary~\cite{llvmvplan,llvmvplanupdate}
  \item VPlan Future Roadmap~\cite{llvmvplan,llvmvplanupdate} (e.g.: combining outer-/inner-loop paths, full function vectorization, etc.)
  \item (Comparison/Outlook auto-vectorization and explicit vectorization~\cite{10.1155/2021/3264624, 9802745, DBLP:conf/vldb/BensonER23})
\end{enumerate}

>>>>

\begin{abstract}
abstract
\end{abstract}

@@

\begin{enumerate}
  \item Modern CPUs (SIMD-Registers, SIMD ISAs)
  \item SIMD performance gains
  \item Auto-Vectorization to help programmers utilize SIMD functionality out of the box 
  (explicit vectorization as time consuming especially when supporting multiple architectures)
  \item LLVM as popular, platform-agnostic compiler framework~\cite{10.5555/977395.977673}
  \item auto-vectorization complexity $\rightarrow$ wish for well designed, 
  interoperable system (VPlan) $\rightarrow$ transition from Loop Vectorizer + SLP Vectorizer to 
  Vectorization Plan $\rightarrow$ ongoing 8 year long effort started by Intel
\end{enumerate}

@@

\section{LLVM's Vectorization Plan}
\label{sec:vplan}
  Overview~\cite{llvmvec,llvmvplan,llvmextloopvec,llvmintrvplan,llvmvplanstate,llvmvplanrv,llvmouterloop,llvmvplanupdate}
\begin{enumerate}
  \item History
  \begin{itemize}
    \item Short Summary how it was before (LoopVectorize)
    \item Initial refactoring proposition by Intel
    \item Describe rough final finished VPlan architecture.
  \end{itemize}
  \item State of VPlan
  \begin{itemize}
    \item Rough algorithm description (interplay of legacy LV, legacy cost model and new VPlan) (with light code references?)
    \item VPlan structure (HCFG, \texttt{VPRegionBlock}, recipes, etc.) and modelling of IR explained via example similar to~\cite{llvmvplanstate} (What is good example? Too simple $\rightarrow$ doesn't cover enough; 
    Too complex $\rightarrow$ doesn't fit into paper.)
    \item (Outer Loop Vectorization?)
  \end{itemize}
  \item Future of VPlan (what is currently being worked on, what is planned next, what has not been planned)
\end{enumerate}

@@

\section{Related Work}
\label{sec:relatedwork}
\begin{enumerate}
  \item Region Vectorizer (RV)~\cite{llvmvplanrv,10.1145/3192366.3192413}
  \item (Other auto-vectorization algorithms / possible improvements~\cite{10.5555/3615924.3615932,4336219,llvmveccontrolflow,saarlandrv,10.1145/3303117.3306172})
  \item (Auto-Vectorization in GCC~\cite{autovecoptgcc, vecoptgcc})
  \item (Comparison with GCC~\cite{gccllvmveccomp,10.1155/2021/3264624})
  \item (Comparison/Outlook auto-vectorization and explicit vectorization~\cite{10.1155/2021/3264624, 9802745, DBLP:conf/vldb/BensonER23})
  \item (Polyhedral compilation techniques~\cite{llvmpolyhedral})
  \item (LLM-based Vectorization~\cite{taneja2024llmvectorizerllmbasedverifiedloop})
\end{enumerate}

@@

\section{Summary and Future Work}
\label{sec:summary}
VPlan Summary

### 2024-11-23 paper/paper.tex
paper

\begin{abstract}
Lorem ipsum
\end{abstract}

@@

While explicit data dependencies are easier to see, even a simple loop, as in 
Figure~\ref{fig:inner-loop-vec}, could carry an implicit dependency in line 3, if
the \texttt{x} and \texttt{y} point to overlapping memory regions (pointer aliasing).

@@

\subsection{Vectorization Costs}
While it may not be legal to vectorize a given control flow, the vectorized code sometimes 
performs worse than its scalar equivalent.

@@

\subsection{History}
LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
simplification has taken place. As VPlan has not yet been integrated into SLP-Vectorization 
we will only focus on explaining the original LoopVectorize structure.

LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
When starting with a initial scalable loop in IR it would first run a legality phase to check 
whetever and how it is legal to vectorize the loop. This phase would produce artifacts/maps like if 
and what runtime aliasing checks would be required. Then the cost model ran next, creating more 
artifacts based on the scalar IR and legality artifacts. These cost model artifacts would 
associate different vectorization factors with a cost. Lastly the transformation phase would 
vectorize the underlying IR in one go based on all the produced artifacts.

However, this rigid flow of LoopVectorize was a problem. Extending LV to include outer loop 
and nested loop vectorizations would have been very challenging, as LV was designed to flatten a
loops control flow into a single basic block. Producing and considering more and 
more artifacts when trying to transform the scalar loop into a vectorized one all at once did also 
not scale well.

As a potential solution Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
using a system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.

The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would use VPlans to represent different 
vectorization approaches, abstracted away from the underlying IR. Each VPlan would be able to 
calculate its cost and transform itself to modify the final IR. At first the Legality 
phase (yellow) would check if the scalar code would be legal to vectorize at all and if so initial 
VPlans would be constructed. In the following Planning phase (magenta) these initial VPlans would 
be optimized using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
Those transformations would be supervised 
by the VPlan's cost model. Finally, the best VPlan would be chosen and materialized/executed, 
modifying the underlying IR. The best VPlan might also be to not vectorize at all (VF $=1$).

@@

In its current state VPlan is deeply integrated into LV.
When auto-vectorizing, LV starts by checking all the different loops in the IR.
If it finds one, it will try to vectorize it (\texttt{LoopVectorizePass::\allowbreak processLoop}).
Several factors might hinder vectorization for a given loop: 
(1) If the loop was already vectorized before, it won't be vectorized again. 
(2) The Legality phase checks if an how it is legal to vectorize the scalar
loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost
of runtime checks for e.g. pointer aliasing can be too high.

The initial VPlans are constructed based on the legality checks and the cost decisions the legacy
LV cost model made referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}).
After initial VPlan construction various VPlan-to-VPlan transformations are applied to optimize
it (\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).

Then the optimal Vectorization Factor (VF) is calculated based on the set of VPlans and their
representive cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), 
which calculation still involves the legacy LV cost model. 
The best VPlan is picked based on the optimal VF. A VF of 1 means it is not
profitable to vectorize the loop, however, interleaving/unrolling can still end up modifying the
underlying IR.
Afterwards, the Interleave Count (IC) is determined 
(\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).

The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In case of
vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.

@@

Figure~\ref{fig:inner-loop-scalar-ir} shows the scalar IR of the main loop basic blocks \texttt{for.body},
\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
hierarchical control-flow-graph (CFG) of a VPlan modelling a vectorization approach for this 
scalar loop. Other basic blocks like the scalar epilogue loop and it's predication check (is the final
vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
in Appendix~\ref{appendix-figures} for the fully modeled HCFG.

@@

\section{Related Work}
\label{sec:relatedwork}
Lorem ipsum

>>>>

\begin{abstract}
  Modern SIMD processors provide various vector registers and ISAs for programmers to utilize. 
  However, manually vectorizing code can be time-consuming, so compilers look to auto-vectorize 
  code for better performance. LLVM, as a widely used compiler toolchain, could only auto-vectorize 
  basic scalar code. Intel proposed a comprehensive refactoring of the underlying system to enable 
  LLVM's auto-vectorization to support more complex optimizations. The suggested Vectorization Plan 
  (VPlan) architecture would be more modular and scalable. Since then, significant efforts have been 
  made to implement this new system. Today, VPlan is already actively involved in vectorizing 
  innermost loops. Other areas, such as outer loop vectorization, are also in development.
\end{abstract}

@@

While explicit data dependencies are easier to see, a simple loop, as in 
Figure~\ref{fig:inner-loop-vec}, could carry an implicit dependency in line 3 if
the \texttt{x} and \texttt{y} point to overlapping memory regions (pointer aliasing).

@@

\subsection{Vectorization Costs}
While vectorizing a given control flow may not be legal, the vectorized code sometimes performs 
worse than its scalar equivalent.

@@

\subsection{History}
LLVM implements two different vectorizers for auto-vectorization, LoopVectorize (LV)
and SLPVectorize. These vectorizers are a part of LLVM's middle-end and run after module 
simplification has taken place. As VPlan still needs to be integrated into SLP-Vectorization, we 
will only focus on explaining the original LoopVectorize structure.

LoopVectorize was designed to focus only on vectorization of innermost loops~\cite{llvmintrvplan}. 
When starting with an initial scalable loop in IR, it would first run a legality phase to check 
whether  and how it is legal to vectorize the loop. This phase would produce artifacts/maps, such as if 
and what runtime aliasing checks would be required. Then, the cost model ran next, creating more 
artifacts based on the scalar IR and legality artifacts. These cost model artifacts would 
associate different vectorization factors with a cost. Lastly, the transformation phase would vectorize 
the underlying IR in one go based on all the produced artifacts.

However, this rigid flow of LoopVectorize presented a challenge. Extending LV to include outer loop 
and nested loop vectorizations would have been difficult, as LV was designed to flatten a
loop's control flow into a single basic block. Producing and considering more and 
more artifacts when trying to transform the scalar loop into a vectorized one all at once also did 
not scale well.

As a potential solution, Intel proposed a long-term effort to refactor LLVM's whole auto-vectorization
using a system called Vectorization Plan (VPlan)~\cite{llvmextloopvec}.

The goal was to unite any auto-vectorization systems (SLP, Inner-/Outer-Loop) into a single flexible
system~\cite{llvmintrvplan,llvmvplanstate}, as seen in Figure~\ref{fig:vplan-future}. This system would 
use VPlans to represent different 
vectorization approaches abstracted away from the underlying IR. Each VPlan would 
calculate its cost and transform itself to modify the final IR. At first, the Legality phase (yellow) 
would check if the scalar code would be legal to vectorize at all, and if so, initial VPlans would 
be constructed. In the following Planning phase (magenta), these initial VPlans would be optimized 
using various VPlan-to-VPlan transformations~\cite{llvmouterloopstatus}. 
The VPlan's cost model would supervise those transformations. Finally, the best VPlan would be chosen 
and materialized/executed, modifying the underlying IR.

@@

In its current state, VPlan is deeply integrated into LV.
When auto-vectorizing, LV starts by checking all the different loops in the IR.
If it finds one, it will try to vectorize it (\texttt{LoopVectorizePass::\allowbreak processLoop}).
Several factors might hinder vectorization for a given loop: 
(1) If the loop was already vectorized before, it will not be vectorized again. 
(2) The Legality phase checks if and how it is legal to vectorize the scalar
loop. (3) The loop's trip count might be too small to make vectorization profitable. (4) The cost 
of runtime checks, such as pointer aliasing, can be too high.

The initial VPlans are constructed based on the legality checks and the legacy LV cost model's 
cost decisions referencing the scalar IR (\texttt{LoopVectorizationPlanner::\allowbreak plan}). 
After initial VPlan construction, various VPlan-to-VPlan 
transformations are applied to optimize it 
(\texttt{LoopVectorizationPlanner::\allowbreak buildVPlansWithVPRecipes}).

Then, the optimal vectorization factor (VF) is calculated based on the set of VPlans and their 
representative cost (\texttt{LoopVectorizationPlanner::\allowbreak cost}), the calculation of 
which still involves the legacy LV cost model.
The best VPlan is picked based on the optimal VF. A VF of 1 means it is not profitable to vectorize 
the loop; however, interleaving/unrolling can still modify the underlying IR. Afterwards, the 
Interleave Count (IC) is determined 
(\texttt{LoopVectorizationCostModel::\allowbreak selectInterleaveCount}).

The best VPlan is then executed (\texttt{LoopVectorizationPlanner::\allowbreak executePlan}). In the 
case of vectorization (VF $>1$), LV will also directly try to vectorize the epilogue loop.

@@

Figure~\ref{fig:inner-loop-scalar-ir} shows the scalar IR of the main loop basic blocks \texttt{for.body},
\texttt{if.then}, and \texttt{for.inc}. Figure~\ref{fig:inner-loop-vplan-hcfg-body} shows the equivalent
hierarchical control-flow-graph (CFG) of a VPlan modeling a vectorization approach for this 
scalar loop. Other basic blocks like the scalar epilogue loop and its predication check (is the final
vector trip count equal to the original trip count) are omitted. See Figure~\ref{fig:inner-loop-vplan-hcfg}
in Appendix~\ref{appendix-figures} for the fully modeled HCFG.

@@

\section{Related Work}
\label{sec:relatedwork}
Besides the main llvm-project's upstream there are many different projects making use of or extending
LLVM's toolchain.
